Associations between chlorophyll *a* and various microcystin health advisory

### 2016-05-25 manuscript/manuscript.Rmd
more edits!

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Microsystin Chlorophyll Manusript}
-->

@@

knit_hooks$set(tab.cap = function(before, options, envir) {
                  if(!before) { 
                    paste('\n\n:', options$tab.cap, sep='') 
                  }
                })
default_output_hook = knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  if (is.null(options$tab.cap) == FALSE) {
    x
  } else
    default_output_hook(x,options)
})
```

@@

#mc_chla[4] <- get_chla(who_rec_low1_cp,0.5)
epa_child <- NULL
who_drink <- NULL
epa_adult <- NULL
who_rec <- NULL
cps<- seq(0.1,0.9,0.1)
for(i in 1:9){
  epa_child[i] <- get_chla(epa_child_cp,cps[i])[1]
  who_drink[i] <- get_chla(who_drink_cp,cps[i])[1]
  epa_adult[i] <- get_chla(epa_adult_cp,cps[i])[1]
  who_rec[i] <- get_chla(who_rec_low1_cp,cps[i])[1]
}
mc_chla<-cbind(epa_child,who_drink,epa_adult,who_rec)

#Agreement Tables
child_conmat<-conmat_to_df(table(chla=nla_dat$CHLA>=mc_chla[5,1],
                                 mcsyt=nla_dat$MCYST_TL_UGL>=0.3))
child_conmat$tots<-child_conmat[,1]+child_conmat[,2]
child_conmat[3,]<-child_conmat[1,]+child_conmat[2,]

@@

total_accuracy<-NULL
total_accuracy$epa_child<-round((child_conmat[1,1]+child_conmat[2,2])/
  sum(child_conmat),2)
total_accuracy$who_drink<-round((who_drink_conmat[1,1]+who_drink_conmat[2,2])/
  sum(who_drink_conmat),2)
total_accuracy$epa_adult<-round((adult_conmat[1,1]+adult_conmat[2,2])/
  sum(adult_conmat),2)
total_accuracy$who_rec<-round((who_rec_conmat[1,1]+who_rec_conmat[2,2])/
  sum(who_rec_conmat),2)

@@

Fortunately, microcystin-LR has been shown to be associated with several other, more commonly measured and well understood components of water quality that are readily assessed in the field.  For instance, there are small or hand held fluorometers that measure chlorohpyll *a*.  Additionally, chlorophyll *a* is a very commonly measured component of water quality that is also known to be positively associated with microsystin-LR concentrations [@pip2014microcystin; @yuan2014managing].  Yuan et. al [-@yuan2014managing] explore these associations in detail and control for other related variables.  In their analysis they find that total nitrogen and chlorophyll *a* show the strongest association with microcystin. Furthermore, they identify chlorophyll*a* and total nitrogen concentrations that are associated with exceeding 1 $\mu$g/L of microcystin.   Given these facts, it should be possible to identify chlorophyll *a* concentrations that would be associated with the new USEPA microcystin health advisory levels for drinking water.  Identifying these associations would provide another tool for water resource managers to help manage the threat to public health posed by cHABs and would be especially useful in the absence of measured microcystin concentrations.

@@

#Results
In the 2007 NLA, microcystin-LR concentrations ranged from `r min(nla_dat$MCYST_TL_UGL,na.rm=T)` to `r max(nla_dat$MCYST_TL_UGL, na.rm=T)` $\mu$g/L.  Microcystin-LR concentrations of 0.05 $\mu$g/L represent the detection limits.  Any value greater than that indicates the presence of microcystin-LR.  Of those lakes with microcystin, the median concentration was `r round(median(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>0.05],na.rm=T), 2)`$\mu$g/L and the mean was `r round(mean(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>0.05],na.rm=T), 2)`$\mu$g/L.  Of all lakes sampled, `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=0.3])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the USEPA Child level, `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=1.6])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the USEPA Adult level, `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=1])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the WHO Drinking level, and `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=2])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the WHO Recreational level. For chlorophyll *a*, the range was `r min(nla_dat$CHLA,na.rm=T)` to `r max(nla_dat$CHLA, na.rm=T)` $\mu$g/L.  All lakes had reported chlorophyll *a* concentrations that exceeded detection limits.  The median concentration was `r median(nla_dat$CHLA,na.rm=T)` $\mu$g/L and the mean was `r round(mean(nla_dat$CHLA,na.rm=T),2)` $\mu$g/L.  The associations between chlorophyll *a* and the upper confidence interval across a range of conditional probability values are shown in Table \ref{tab:mc_chla_table}.  Specific chlorophyll *a* that are associated with greater than even odds of exceeding the advisory levels were `r mc_chla[1]`, `r mc_chla[2]`, `r mc_chla[3]`, and `r mc_chla[4]`$\mu$g/L for 0.3, 1.0, 1.6, and 2.0 $\mu$g/L advisory levels, respectively (Table \ref{tab:mc_chla_table} & Figure \ref{fig:multi_cp_plot}).

The chlorophyll *a* cutoffs may be used to predict whether or not a lake exceeds the microcystin health advisories.  Doing so allows us to compare the accuracy of the prediction as well as evaluate false negatives.  Total accuracy of the four cutoffs predicting microcystin exceedances were `r total_accuracy$epa_child*100`% for the USEPA children's advisory, `r total_accuracy$who_drink*100`% for the WHO drinking water advisory,  `r total_accuracy$epa_adult*100`% for the USEPA adult advisory, and `r total_accuracy$who_rec*100`% for the WHO regreational advisory (Tables \ref{tab:child_conmat_table}, \ref{tab:who_drink_conmat_table}, \ref{tab:adult_conmat_table}, & \ref{tab:who_rec_conmat_table}).  However, total accuracy is only one part of the prediction performace with which we are concerned.

@@

#Discussion
The association between log microcystin-LR and log chlorophyll *a* shows a wedge pattern (Figure \ref{fig:chla_micro_scatter}).  This indicates that, in general, higher concentrations of microcystin-LR almost always co-occur with higher concentrations of chlorophyll *a* yet the inverse is not true.  Higher chlorophyll *a* is not necessarily predictive of higher microcystin-LR concentrations; however, chlorophyll *a* may be predictive of the probability of exceeding a certain threshold.

@@

A second potential use is to identify past bloom events from historical data.  As harmful algal blooms are made up of many species and have various mechanisms responsible for adverse impacts (e.g., toxins, hypoxia, odors), there is no single definition of a bloom.  For cHABs, one approach has been to identify an increase over a baseline concentration of phycocyanin [@miller2013spatiotemporal].  This is a useful approach for targeted studies, but phycocyanin is also not always available and measures the predominance of cyanobacterial pigments and not toxins. Using our chlorophyll *a* cutoffs provides a value that is more directly associated with microcystin-LR and can be used to classify lakes, from past surveys, as having bloomed.

The values we propose are national and may miss regional variation in water quality, including, chlorophyll *a* and microcystin-LR [@beaver2014land, @wagner2011landscape]. A set of regional conditional probabilities would be interesting; however, limiting the analysis to the data available per region would make interpretaion difficult.  The sample size for each of the conditional probabilities would be reduced (it ranges from `r  min(table(nla_dat$WSA_ECO9))` to `r  max(table(nla_dat$WSA_ECO9))`) and the number of lakes in each region that exceed the microcystin values is also reduced.  The result is that our confidence in the conditional probabilites would be less (i.e. greatly increased confidence intervals) and the relationships less pronounced as we have fewer lakes on which to base the probabilites.  Thus, this dataset is best for making national scale recommendations.

>>>>

<!--marion2012vivo
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Microsystin Chlorophyll Manusript}
-->

@@

knit_hooks$set(tab.cap = function(before, options, envir) {
                  if(!before) { 
                    paste('\n\n:', options$tab.cap, sep='') 
                  }
                })https://github.com/USEPA/Microcystinchla/blob/master/R/get_nla.R
default_output_hook = knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  if (is.null(options$tab.cap) == FALSE) {
    x
  } else
    default_output_hook(x,options)
})
```

@@

#mc_chla[4] <- get_chla(who_rec_low1_cp,0.5)
epa_child <- NULL
who_drink <- NULL
epa_adult <- NULL
who_rec <- NULL
cps<- seq(0.1,0.9,0.1)
for(i in 1:9){
  epa_child[i] <- get_chla(epa_child_cp,cps[i])[1]
  who_drink[i] <- get_chla(who_drink_cp,cps[i])[1]
  epa_adult[i] <- get_chla(epa_adult_cp,cps[i])[1]
  who_rec[i] <- get_chla(who_rec_low1_cp,cps[i])[1]
}
mc_chla<-cbind(epa_child,who_drink,epa_adult,who_rec)
mc_chla_rnd <- mc_chla[mc_chla>0.1]
#Agreement Tables
child_conmat<-conmat_to_df(table(chla=nla_dat$CHLA>=mc_chla[5,1],
                                 mcsyt=nla_dat$MCYST_TL_UGL>=0.3))
child_conmat$tots<-child_conmat[,1]+child_conmat[,2]
child_conmat[3,]<-child_conmat[1,]+child_conmat[2,]

@@

total_accuracy<-NULL
total_accuracy$epa_child<-https://github.com/USEPA/Microcystinchla/blob/master/R/get_nla.R((child_conmat[1,1]+child_conmat[2,2])/
  sum(child_conmat),2)
total_accuracy$who_drink<-round((who_drink_conmat[1,1]+who_drink_conmat[2,2])/
  sum(who_drink_conmat),2)
total_accuracy$epa_adult<-round((adult_conmat[1,1]+adult_conmat[2,2])/
  sum(adult_conmat),2)
total_accuracy$who_rec<-round((who_rec_conmat[1,1]+who_rec_conmat[2,2])/
  sum(who_rec_conmat),2)

@@

Fortunately, cyanobacteria and microcystin-LR has been shown to be associated with several other, more commonly measured and well understood components of water quality that are readily assessed in the field [e.g. @yuan2015deriving].  For instance, there are small or hand held fluorometers that measure chlorohpyll *a*.  Additionally, chlorophyll *a* is a very commonly measured component of water quality that is also known to be positively associated with microsystin-LR concentrations [@pip2014microcystin; @yuan2014managing].  Yuan et. al [-@yuan2014managing] explore these associations in detail and control for other related variables.  In their analysis they find that total nitrogen and chlorophyll *a* show the strongest association with microcystin. Furthermore, they identify chlorophyll*a* and total nitrogen concentrations that are associated with exceeding 1 $\mu$g/L of microcystin.   Given these facts, it should be possible to identify chlorophyll *a* concentrations that would be associated with the new USEPA microcystin health advisory levels for drinking water.  Identifying these associations would provide another tool for water resource managers to help manage the threat to public health posed by cHABs and would be especially useful in the absence of measured microcystin concentrations.

@@

#Results
In the 2007 NLA, microcystin-LR concentrations ranged from `r min(nla_dat$MCYST_TL_UGL,na.rm=T)` to `r max(nla_dat$MCYST_TL_UGL, na.rm=T)` $\mu$g/L. Microcystin-LR concentrations of 0.05 $\mu$g/L represent the detection limits.  Any value greater than that indicates the presence of microcystin-LR.  Of those lakes with microcystin, the median concentration was `r round(median(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>0.05],na.rm=T), 2)`$\mu$g/L and the mean was `r round(mean(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>0.05],na.rm=T), 2)`$\mu$g/L.  Of all lakes sampled, `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=0.3])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the USEPA Child level, `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=1.6])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the USEPA Adult level, `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=1])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the WHO Drinking level, and `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=2])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the WHO Recreational level. For chlorophyll *a*, the range was `r min(nla_dat$CHLA,na.rm=T)` to `r max(nla_dat$CHLA, na.rm=T)` $\mu$g/L.  All lakes had detectable levels of chlorophyll *a*.  The median concentration was `r median(nla_dat$CHLA,na.rm=T)` $\mu$g/L and the mean was `r round(mean(nla_dat$CHLA,na.rm=T),2)` $\mu$g/L.  The associations between chlorophyll *a* and the upper confidence interval across a range of conditional probability values are shown in Table \ref{tab:mc_chla_table}.  Specific chlorophyll *a* that are associated with greater than even odds of exceeding the advisory levels were `r mc_chla[1]`, `r mc_chla[2]`, `r mc_chla[3]`, and `r mc_chla[4]`$\mu$g/L for 0.3, 1.0, 1.6, and 2.0 $\mu$g/L advisory levels, respectively (Table \ref{tab:mc_chla_table} & Figure \ref{fig:multi_cp_plot}).

The chlorophyll *a* cutoffs may be used to predict whether or not a lake exceeds the microcystin health advisories.  Doing so allows us to compare the accuracy of the prediction as well as evaluate false negatives.  Total accuracy of the four cutoffs predicting microcystin exceedances were `r total_accuracy$epa_child*100`% for the USEPA children's drinking water advisory, `r total_accuracy$who_drink*100`% for the WHO drinking water advisory,  `r total_accuracy$epa_adult*100`% for the USEPA adult drinking water advisory, and `r total_accuracy$who_rec*100`% for the WHO regreational advisory (Tables \ref{tab:child_conmat_table}, \ref{tab:who_drink_conmat_table}, \ref{tab:adult_conmat_table}, & \ref{tab:who_rec_conmat_table}).  However, total accuracy is only one part of the prediction performace with which we are concerned.

@@

#Discussion
The log-log association between microcystin-LR and chlorophyll *a* indicates that, in general, higher concentrations of microcystin-LR almost always co-occur with higher concentrations of chlorophyll *a* yet the inverse is not true (Figure \ref{fig:chla_micro_scatter}).  Higher chlorophyll *a* is not necessarily predictive of higher microcystin-LR concentrations; however, chlorophyll *a* may be predictive of the probability of exceeding a certain threshold.

@@

A second potential use is to identify past bloom events from historical data.  As harmful algal blooms are made up of many species and have various mechanisms responsible for adverse impacts (e.g., toxins, hypoxia, odors), there is no single definition of a bloom.  For cHABs, one approach has been to utilize phycocyanin to screen for or identify bloom events [@miller2013spatiotemporal, ahn2007alternative, @marion2012vivo].  This is a useful approach, but phycocyanin is not always available, thus limiting its utility especially for examining historical data. Using our chlorophyll *a* cutoffs provides a value that is also associated with microcystin-LR and can be used to classify lakes, from past surveys, as having bloomed.

The values we propose are national and may miss regional variation in water quality, including, chlorophyll *a* and microcystin-LR [@beaver2014land, @wagner2011landscape]. A set of regional conditional probabilities would be interesting; however, limiting the analysis to the data available per region would make interpretaion difficult.  The sample size for each of the conditional probabilities would be reduced (it ranges from `r  min(table(nla_dat$WSA_EChttps://github.com/USEPA/Microcystinchla/blob/master/R/get_nla.RO9))` to `r  max(table(nla_dat$WSA_ECO9))`) and the number of lakes in each region that exceed the microcystin values is also reduced.  The result is that our confidence in the conditional probabilites would be less (i.e. greatly increased confidence intervals) and the relationships less pronounced as we have fewer lakes on which to base the probabilites.  Thus, this dataset is best for making national scale recommendations.

Two other limitations with the 2007 National Lakes Assessment data are that they represent a single sample from that lake and do not capture temporal dynamics and without subsetting the data do not provide us the ability to validate the presence of microcystin-LR.  As of this writing, the 2012 National Lakes Assessment data are not public.  When these data are released, a validation of this approach can be completed then.
